---
output: html_document
---
====================================================================================================

# Reproducible Research Assignment 1

====================================================================================================

First, set the working directory for the project and load all necessary libraries
```{r, echo=FALSE}
setwd("C:/users/celia/Coursera/ReproducibleResearch/Assignment1")
```

```{r, echo=TRUE}
library("plyr")
library("reshape2")
library("datasets")
library("ggplot2")
```

## Loading and preprocessing the data
    
    1. Check if the source file already exists, if not, download/unzip dataset
    
```{r}
fileName <- "activity.csv"
if (!file.exists(fileName)) {
    
    url1<- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(url1,"./repdata_data_activity")
    unzip("./repdata_data_activity")
    
}
```

    2. Load the dataset and process the data infor a format suitable for analysis

```{r}
activity <- read.csv(fileName, colClasses=c("numeric","Date", "character"))
```


## What is mean total number of steps taken per day? 
    
    1. Create dataframe by summarizing total steps by day
    
```{r}
daily_tot_steps <- ddply(activity, .(date), summarize, sum(steps, na.rm = TRUE))
colnames(daily_tot_steps) <- c("date", "total_steps")
```
    
    2. Plot histogram of the total number of steps taken each day

#### Histogram of the total number of steps take each day 

```{r}
hist(daily_tot_steps$total_steps, main="Total Steps Taken per Day", col="blue",
     xlab="Total Steps", plot = TRUE)
```

    3. Calculate the Mean and Median total number of steps taken per day

```{r, results='markup'}
mean_total_per_day <- mean(daily_tot_steps$total_steps, na.rm=TRUE)
mean_total_per_day
```

#### The mean total number of steps taken per day is `r mean_total_per_day` .

```{r, results='markup'}
median_total_per_day <- median(daily_tot_steps$total_steps, na.rm=TRUE)
median_total_per_day
```

#### The median total number of steps take per day is `r median_total_per_day` .

## What is the average daily pattern?

    1. Calculate total, mean, and median by date
    
```{r}
daily_sum <- ddply(activity, .(date), function(x)
                                    c(daily_total=sum(x$steps,na.rm=TRUE), 
                                    daily_mean=mean(x$steps, na.rm=TRUE), 
                                    daily_median=median(x$steps, na.rm=TRUE)))
```
    
    2. Calculate total, mean, and median by interval across all days 
    
```{r}
interval_sum <- ddply(activity, .(interval), function(x)
                                    c(interval_total=sum(x$steps,na.rm=TRUE), 
                                    interval_mean=mean(x$steps, na.rm=TRUE), 
                                    interval_median=median(x$steps, na.rm=TRUE)))

order_interval_sum <- interval_sum[order(as.numeric(interval_sum[,1]), decreasing=FALSE),]

max_interval <- head(interval_sum[order(as.numeric(interval_sum[,2]), decreasing=TRUE), ], 1)
max_interval
```

    3. Time series plot of 5-minute interval (x) and the average number of steps taken 

#### Time series plot of the 5-minute interval and the average number of steps taken
```{r}
with(order_interval_sum, plot(interval, interval_mean, main="Average number of steps taken at each interval",
                              type="l", col="black", xlab="Interval", ylab="Average Steps Taken"))
```

#### The interval for the maximum number of average steps taken across all days is `r max_interval$interval`.


## Imputting missing values
    
    1. count the number of missing values the dataset
    
```{r}
ms <- which(is.na(activity$steps))
nm <- length(ms)

```

#### The total number of missing values (NA) in the dataset is `r nm`. 

    2. Create dataset with the missing steps value filled in with the mean of each interval across all days
    
```{r}
activity_clean <- transform(activity, 
          steps = ifelse(is.na(steps), 
                          ave(steps, interval, FUN = function(x) mean(x, na.rm = TRUE)), 
                          steps))
nrow(activity_clean)
``` 

    3. Create dataframe by summarizing total steps by day
    
```{r}
new_daily_total <- ddply(activity_clean, .(date), summarize, sum(steps, na.rm = FALSE))
colnames(new_daily_total) <- c("date", "total_steps")
```

    4. Plot the histogram with daily total step taken with the new dataset

#### Histogram of the new daily total steps taken after imputting the missing numbers

```{r}
hist(new_daily_total$total_steps, main="Total Steps Taken per Day", col="green",
     xlab="Total Steps", plot = TRUE)
```

    5. Mean and median total number of steps taken per day
    
```{r}
new_mean_total_per_day <- mean(new_daily_total$total_steps, na.rm=TRUE)
new_mean_total_per_day
new_median_total_per_day <- median(new_daily_total$total_steps, na.rm=TRUE)
new_median_total_per_day
```

#### The new dataset behaves differently from the original dataset.
#### The new mean total number of steps taken per day is `r new_mean_total_per_day` .

#### The new median total number of steps taken per day is `r new_median_total_per_day` .


## Are there differences in activity patterns between weekdays and weekends?

    1. Add a column to indicate day of the week for each date and determin if weekday or weekend
    
```{r}
activity_clean$dayofweek <- weekdays(activity_clean$date)

activity_clean$wd_we <- ifelse(!activity_clean$dayofweek %in% c("Saturday", "Sunday"),"weekday","weekend")

```

    2. create a new dataframe with weekend only activity and calculate the average steps for each interval
    
```{r}
weekend_activity <- subset(activity_clean, (activity_clean$wd_we == "weekend"), select=names(activity_clean))
weekend_interval_sum <- ddply(weekend_activity, .(interval), function(x)
                            c(weekend_interval_total=sum(x$steps,na.rm=TRUE), 
                                weekend_interval_mean=mean(x$steps, na.rm=TRUE), 
                                weekend_interval_median=median(x$steps, na.rm=TRUE)))

order_weekend_interval_sum <- weekend_interval_sum[order(as.numeric(weekend_interval_sum[,1]), decreasing=FALSE),]
```

    3. create a new dataframe with weekday only activity and calculate the average steps for each interval

```{r}
weekday_activity <- subset(activity_clean, (activity_clean$wd_we == "weekday"), select=names(activity_clean))
weekday_interval_sum <- ddply(weekday_activity, .(interval), function(x)
                            c(weekday_interval_total=sum(x$steps,na.rm=TRUE), 
                                weekday_interval_mean=mean(x$steps, na.rm=TRUE), 
                                weekday_interval_median=median(x$steps, na.rm=TRUE)))

order_weekday_interval_sum <- weekday_interval_sum[order(as.numeric(weekday_interval_sum[,1]), decreasing=FALSE),]

```

    4. Plot the average number of steps taken per day for each interval for weekend and weekdays

#### Panel plot containing a time series plot of the 5-minute interval and average steps taken per day for weekend and anotherfor weekday.

```{r}
par(mfrow=c(2,1))

with(order_weekend_interval_sum, plot(interval, weekend_interval_mean, main="Weekend",
                              type="l", col="blue", xlab="Interval", ylab="Average Steps Taken"))

with(order_weekday_interval_sum, plot(interval, weekday_interval_mean, main="Weekday",
                              type="l", col="red", xlab="Interval", ylab="Average Steps Taken"))

```